---
title: "Homework 1 - Exploring Environmental Injustices"
author: "Michaela Dennis"
format: pdf
execute:
  warning: false
  message: false
---

## Load packages and read in data

```{r}
#| warning: false
#| message: false
library(tidyverse)
library(sf)
library(tmap)

# read in geodatabase of EJScreen data at the Census Block Group level
ejscreen <- sf::st_read(
  'data/ejscreen/EJSCREEN_2023_BG_StatePct_with_AS_CNMI_GU_VI.gdb') 

# filter to a state you are interested in
maryland <- ejscreen %>%
  dplyr::filter(ST_ABBREV == "MD") 

# find the average values for all variables within counties
maryland_counties <- aggregate(maryland,
                              by = list(maryland$CNTY_NAME),
                              FUN = mean, na.rm = TRUE)
                        
# filter to a county you are interested in
baltimore <- maryland_counties %>%
  dplyr::filter(Group.1 %in% c("Baltimore City"))
pg <- maryland_counties %>%
  dplyr::filter(Group.1 %in% c("Prince George's County"))
```


## Visualize
### Plotting the percentage of people of color in each Maryland county
```{r}
tm_shape(maryland_counties) + 
  tm_graticules() +
  tm_polygons(fill = "P_PEOPCOLORPCT", # People of color (%)
              fill.scale = tm_scale_continuous(
                values = "brewer.blues"), # create a color scale
              fill.legend = tm_legend(
                title = "\nPeople of\ncolor (%)\n", # break in 
                # the legend title so the legend doesn't look so wide
                                      position = tm_pos_out(
                                        "right", 
                                        "center"), # don't cover map
                                      title.size = 0.7,
                                      text.size = 0.5,
                                      height = 17)) +
tm_shape(baltimore, layer = "Shape") +
  tm_borders(col = "red") +
  tm_shape(pg, layer = "Shape") +
    tm_borders(col = "red") +
  tm_title(
    "People of Color by County, Maryland (EJScreen 2023)") +
  tm_compass(
    position = tm_pos_in(
      "left",
       "bottom"))   # make sure the compass doesn't cover data

  
```

This map depicts the percentage of people of color in each county in Maryland. As demonstrated by this map, Baltimore City and Prince George's County have the highest percentages of people of color. Black people make up almost 60% of the population in both of these counties ([Neilsberg](https://www.neilsberg.com/insights/lists/counties-in-maryland-by-black-population/)). Both of these counties also happen to be located in or next to two major cities, Baltimore and Washington D.C..


### Plotting Diesel particulate matter
```{r}
tm_shape(maryland_counties) + 
  tm_graticules() +
  tm_polygons(fill = "DSLPM", # diesel pm in the air by county
              fill.scale = tm_scale_continuous(
                values = "PurpOr"), # create a color scale
              fill.legend = tm_legend(
                title = "\nDiesel PM (µg/m^3)\n\n", 
                                      position = tm_pos_out(
                                        "right",
                                         "center"), # don't cover map
                                      title.size = 0.7,
                                      text.size = 0.5,
                                      height = 16)) +
tm_shape(baltimore, layer = "Shape") +
  tm_borders(col = "red") +
  tm_shape(pg, layer = "Shape") +
    tm_borders(col = "red") +
  tm_title("Diesel Particulate Matter by County, Maryland (EJScreen 2023)") +
  tm_compass(position = tm_pos_in(
    "left",
     "bottom"))   # make sure the compass doesn't cover data

```

This map showcases the average diesel particulate matter (PM) exposure a person may encounter in each county. Diesel particulate matter is measured in µg/m^3. Diesel PM mostly comes from exhaust ([EPA](https://web.archive.org/web/20250123161322/https://www.epa.gov/ejscreen/ejscreen-map-descriptions#ejin)). We can see that diesel PM is highest in counties that also have the highest percentages of people of color.


### Plotting the air toxics cancer risk

```{r}
tm_shape(maryland_counties) + 
  tm_graticules() +
  tm_polygons(fill = "CANCER", # air toxics inhalation cancer risk
              fill.scale = tm_scale_continuous(
                values = "SunsetDark"), # create a color scale
              fill.legend = tm_legend(
                title = "\nLifetime cancer\nrisk (per million)\n",
                                      position = tm_pos_out(
                                        "right",
                                         "center"),
                                      title.size = 0.7,
                                      text.size = 0.5,
                                      height = 15)) +
tm_shape(baltimore, layer = "Shape") +
  tm_borders(col = "red") +
tm_shape(pg, layer = "Shape") +
  tm_borders(col = "red") +  
  tm_title("Air Toxics Cancer Risk by County, Maryland (EJScreen 2023)") +
  tm_compass(position = tm_pos_in(
    "left",
     "bottom"))   # make sure the compass doesn't cover any of the data


```

Lastly, this third map shows the air toxics cancer risk. The data shows the estimated lifetime inhalation cancer risk from carcinogens in outdoor air from the EPA’s AirToxScreen ([EPA](https://web.archive.org/web/20250123161322/https://www.epa.gov/AirToxScreen)). The measure of this data is persons per million over the span of 70 years (estimated one lifetime). This map serves the purpose of tying the first two maps together. People of color throughout the U.S. often face disproportionate environmental risks and thus experience higher rates of health complications and shorter life spans. Although these data may not necessarily be causal, it is important to note the patterns.


## Citations
### Data
EJScreen: Environmental Justice Screening and Mapping Tool. 
PEDP, pedp-ejscreen.azurewebsites.net/. Accessed 4 Oct. 2026. 


### References
EJ Screen. 
“Ejscreen Map Descriptions.” US EPA, 19 Dec. 2024,\
web.archive.org/web/20250123161322/https://www.epa.gov/\
ejscreen/ejscreen-map-descriptions#ejin. 

Neilsberg. 
Counties in Maryland Ranked by Black Population - 2025\
| Neilsberg, www.neilsberg.com/insights/lists/\
counties-in-maryland-by-black-population/. Accessed 4 Oct. 2026. 

EPA. 
“Air Toxics Screening Assessment.” US EPA, 7 Jan. 2025,\
web.archive.org/web/20250123161049/\
https://www.epa.gov/AirToxScreen. 