# Homework 1 - Exploring environmental injustices
Read in data

```{r}
#| warning: false
#| message: false
library(tidyverse)
library(sf)
library(here)
library(tmap)

# read in geodatabase of EJScreen data at the Census Block Group level
ejscreen <- sf::st_read('data/ejscreen/EJSCREEN_2023_BG_StatePct_with_AS_CNMI_GU_VI.gdb') 

# filter to a state you are interested in
maryland <- ejscreen %>%
  dplyr::filter(ST_ABBREV == "MD") 

# filter to a county you are interested in
baltimore <- ejscreen %>%
  dplyr::filter(CNTY_NAME %in% c("Baltimore City"))

# find the average values for all variables within counties
maryland_counties <- aggregate(maryland, by = list(maryland$CNTY_NAME), FUN = mean, na.rm = TRUE)

```


### Plotting the percentage of people of color in each Maryland county.
```{r}
tm_shape(maryland_counties) + 
  tm_graticules() +
  tm_polygons(fill = "P_PEOPCOLORPCT", # % POC by county avg
              fill.scale = tm_scale_continuous(values = "brewer.blues"), # create a color scale
              fill.legend = tm_legend(title = "\nPercentage of\nPeople of Color\n", # break in the legend title so the legend doesn't look so wide
                                      position = tm_pos_out("right", "center"), # make sure it doesn't cover the map
                                      title.size = 0.7,
                                      text.size = 0.5,
                                      height = 17)) +
  tm_title("Percentage of People of Color") +
  tm_compass(position = tm_pos_in("left", "bottom"))   # make sure the compass doesn't cover any of the data

```

This map depicts the percentage of people of color in each county in Maryland. As demonstrated by this map, Baltimore City and Prince George's County have the highest percentages of Black people. Black people make up almost 60% of the population in both of these counties ([nielsberg.com]('https://www.neilsberg.com/insights/lists/counties-in-maryland-by-black-population/')). Both of these counties also happen to be located in or next to two major cities, Baltimore and Washington D.C..


### Plotting Diesel particulate matter.
```{r}
tm_shape(maryland_counties) + 
  tm_graticules() +
  tm_polygons(fill = "DSLPM", # % POC by county avg
              fill.scale = tm_scale_continuous(values = "PurpOr"), # create a color scale
              fill.legend = tm_legend(title = "\nDiesel Particulate\nMatter (µg/m^3)\n\n", # break in the legend title so the legend doesn't look so wide
                                      position = tm_pos_out("right", "center"), # make sure it doesn't cover the map
                                      title.size = 0.7,
                                      text.size = 0.5,
                                      height = 16)) +
  tm_title("Diesel Particulate Matter") +
  tm_compass(position = tm_pos_in("left", "bottom"))  # make sure the compass doesn't cover any of the data

```

This map showcases the average diesel particuplate matter (PM) exposure a person may encounter in each county. Diesel particulate matter is measured in µg/m^3. Diesel PM mostly comes from exhaust ([EPA Air Toxic's Screening]('https://web.archive.org/web/20250123161322/https://www.epa.gov/ejscreen/ejscreen-map-descriptions#ejin')). We can see that diesel PM is highest in counties that also have the highest perentages of Black residents.


### Plotting the air toxics cancer risk.

```{r}
tm_shape(maryland_counties) + 
  tm_graticules() +
  tm_polygons(fill = "CANCER", # % POC by county avg
              fill.scale = tm_scale_continuous(values = "SunsetDark"), # create a color scale
              fill.legend = tm_legend(title = "\nAir Toxics Cancer Risk\n", # break in the legend title so the legend doesn't look so wide
                                      position = tm_pos_out("right", "center"), # make sure it doesn't cover the map
                                      title.size = 0.7,
                                      text.size = 0.5,
                                      height = 15)) +
  tm_title("Air Toxics Cancer Risk") +
  tm_compass(position = tm_pos_in("left", "bottom"))  # make sure the compass doesn't cover any of the data

```

Lastly, this third map shows the air toxics cancer risk. The data shows the estimated lifetime inhalation cancer risk from carcinogens in outdoor air from the EPA’s AirToxScreen ([EPA]('https://web.archive.org/web/20250123161322/https://www.epa.gov/AirToxScreen')). The measure of this data is persons per million over the span of 70 years (one lifetime). This map serves the purpose of tying the first two maps together. People of color throughout the U.S. often face disproportionate environmental risks and thus experience higher rates of health complications and shorter life spans. Although these data may not necessarily be causal, it is important to note the patterns.
